### 2023-10-03 main.tex
biblio
Hadamard

The way one can define a multi-type (or multi-species) interacting particle system is not unique. In the literature there are many. Two of them are the following. On one hand, the \textit{multi-layer models}, in which the process is constructed on a many layer graph where each graph correspond to a type and perform a kind of dynamic (independent jump, exclusion/inclusion walkers) and the layers communicate each other with some "switching probabilities". The state space of the Markov process is the Cartesian product of the state spaces of each layer, i.e. of the two species. For example see \cite{floreani2022switching,redig2022ergodic,redig2023equilibrium}. On the other hand, the \textit{single layer models}. Here, there is just one layer, where particles of each type can pile up, in a way that the exclusion dynamic concerns many species together. In other words, now the species can interact "directly", not just by switching layer. Here the state space contains all the species and it is not product anymore. Some examples of such a dynamic can be found in \cite{casini2022uphill,vanicat2017exact,zhou2021orthogonal}. \\

@@

% However, they would lead to duality matrices where the number of dual particles is greater than the number of original ones, loosing the simplification of the dynamics of the dual process. 
 \begin{flushright}
     $\square$
 \end{flushright}
 \paragraph{Proof of \eqref{siteDualRelation}.} To prove \eqref{siteDualRelation} we need to transform via the Hadamard formula the transposed of the site Hamiltonian \eqref{siteDualHamiltonian} and then to  introduce properly a creation operator acting on an extra-site $u(x)$. % that we connect to every site $x\in V$ of the graph $G$. \\
 Considering $A,B\in gl(N)$, the Hadamard formula reads 
 \begin{equation}\label{HadamardFormula}
     \exp{(-B)}A\exp{(B)}=A-\left[B,A\right]+\frac{1}{2!}\left[B,\left[B,A\right]\right]-\frac{1}{3!}\left[B,\left[B,\left[B,A\right]\right]\right]+\ldots
 \end{equation}
In the following we need the application of the Hadamard formula with $B=\sum_{a=1}^{N-1}E_{aN}$ defined in \eqref{EquationEx} and with $A$ equal to some of the generator $E_{ab}$ of the Lie algebra. Therefore we compute 
\begin{enumerate}
    \item for $A=E_{b N}$ with $b\in \{1,\ldots,N-1\}$, we obtain 
    \begin{equation}\label{HT_El1}
        \exp{(-E)}E_{b N}\exp{(E)}=E_{b N}
    \end{equation}
    because $E_{b N}$ commutes with $E$
    \item for $A=E_{bb}$ with $b \in \{1,\ldots,N-1\}$, we obtain 
    \begin{equation}\label{HT_Ell}
        \exp{(-E)}E_{b b}\exp{(E)}=E_{b b}+E_{b N}
    \end{equation}
Indeed, using \eqref{eq:comgl} we have 
    \begin{equation}
       \left[E, E_{bb}\right]=\sum_{a=1}^{N-1}\left(E_{ab}\delta_{b N}-E_{b N}\delta_{ab}\right)=-E_{bN}
    \end{equation}
    Inserting the above commutator in \eqref{HadamardFormula}, we obtain \eqref{HT_Ell}.
    \item for $A=E_{Nb}$ with $b\in \{1,\ldots,N-1\}$, we obtain 
    \begin{equation}\label{HT-E1l}
        \exp{(-E)}E_{N a}\exp{(E)}=E_{Na}+E_{NN}-\sum_{c=1}^{N-1}\left(E_{cN}+E_{ca}\right)
    \end{equation}
  %\com{Why indices have names $b, k, j, a, b$? Fix one set for the generators}

    Indeed, using \eqref{eq:comgl} we have 
    \begin{equation}
[E,E_{Na}]=\sum_{c=1}^{N-1}\left(E_{ca}\delta_{NN}-E_{NN}\delta_{ca}\right)= (\sum_{c=1}^{N-1}E_{ca} ) -E_{NN};
\end{equation}
and 
\begin{equation}
\begin{split}
\left[E,[E,E_{Na}]\right]&=\sum_{d=1}^{N-1}\sum_{c=1}^{N-1}\left[E_{dN},E_{ca}\right]-\sum_{d=1}^{N-1}\left[E_{dN},E_{NN}\right]
\\&=
\sum_{d,c=1}^{N-1}\left(E_{da}\delta_{cN}-E_{cN}\delta_{da}\right)-\sum_{d=1}^{N-1}\left(E_{dN}\delta_{NN}-E_{NN}\delta_{dN}\right)
\\=&
-2\sum_{c=1}^{N-1}E_{cN};
\end{split}
\end{equation}
Inserting the above commutators in \eqref{HadamardFormula}, we obtain \eqref{HT-E1l}.
\item for $A=E_{ba}$ with $a,b \in \{1,\ldots,N-1\}$ we obtain 
\begin{equation}\label{HT-Ekl}
    \exp{(-E)}E_{a b}\exp{(E)}=E_{ba}+E_{bN}
\end{equation}
   Indeed, using \eqref{eq:comgl} we have 
\begin{equation}
[E,E_{ba}]=\sum_{c=1}^{N-1}\left(E_{ca}\delta_{b1}-E_{bN}\delta_{c a}\right)=-E_{bN};
\end{equation}
Inserting the above commutator in \eqref{HadamardFormula}, we obtain \eqref{HT-Ekl}.
\item for $A=E_{NN}$ we obtain 
\begin{equation}\label{HT-E11}
    \exp{(-E)}E_{NN}\exp{(E)}=E_{NN}-\sum_{a=1}^{N-1}E_{aN}
\end{equation}
  Indeed, using \eqref{eq:comgl} we have

@@

First, we multiply both sides by $R_{x}\exp{(E^{x})}$

@@

then,we insert the identity $I=\exp{(E^{x})}\exp{(-E^{x})}$ in the right hand side

@@

 Using \eqref{HT_El1}, \eqref{HT_Ell},\eqref{HT-E1l}, \eqref{HT-Ekl}, \eqref{HT-E11} we have that 

>>>>

The way one can define a multi-type (or multi-species) interacting particle system is not unique. In the literature there are many. Two of them are the following. On one hand, the \textit{multi-layer models}, in which the process is constructed on a many layer graph where each graph correspond to a type and perform a kind of dynamic (independent jump, exclusion/inclusion walkers) and the layers communicate each other with some "switching probabilities". The state space of the Markov process is the Cartesian product of the state spaces of each layer, i.e. of the two species. For example see \cite{floreani2022switching,redig2022ergodic,redig2023equilibrium}. On the other hand, the \textit{single layer models}. Here, there is just one layer, where particles of each type can pile up, in a way that the exclusion dynamic concerns many species together. In other words, now the species can interact "directly", not just by switching layer. Here the state space contains all the species and it is not product anymore. Some examples of such a dynamic can be found in \cite{casini2022uphill,vanicat2017exact,zhou2021orthogonal,casini2023density}. \\

@@

% However, they would lead to duality matrices where the number of dual particles is greater than the number of original ones, loosing the simplification of the dynamics of the dual process. 
 \begin{flushright}
     $\square$
 \end{flushright}
 \paragraph{Proof of \eqref{siteDualRelation}.} To prove \eqref{siteDualRelation} we need to transform via the Hadamard formula the transposed of the site Hamiltonian \eqref{siteDualHamiltonian} and then to  introduce properly a creation operator acting on an extra-site $u(x)$. % that we connect to every site $x\in V$ of the graph $G$. \\
 Considering $A,B\in gl(N)$, the Hadamard formula reads 
 \begin{equation}\label{HadamardFormula}
     \exp{(-B)}A\exp{(B)}=A-\left[B,A\right]+\frac{1}{2!}\left[B,\left[B,A\right]\right]-\frac{1}{3!}\left[B,\left[B,\left[B,A\right]\right]\right]+\ldots
 \end{equation}
In the following we need the application of the Hadamard formula with $B=\sum_{a=1}^{N-1}E_{aN}$ defined in \eqref{EquationEx} and with $A$ equal to some of the generator $E_{cd}$ of the Lie algebra with $c,d\in\{1,\ldots,N\}$. Therefore we compute 
\begin{align}
	\left[B,A\right]=\sum_{a=1}^{N-1}\left[E_{aN},E_{cd}\right]=\delta_{cN}\sum_{a=1}^{N-1}E_{ad}-\delta_{d<N}E_{cN}
\end{align}
and 
\begin{align}
	\left[B\left[B,A\right]\right]=&\left[\sum_{a=1}^{N-1}E_{aN},\delta_{cN}\sum_{b=1}^{N-1}E_{bd}-\delta_{d<N}E_{cN}\right]\nonumber\\=&
	\sum_{a=1}^{N-1}\left(\delta_{cN}\sum_{b=1}^{N-1}\left[E_{aN},E_{bd}\right]-\delta_{d<N}\left[E_{aN},E_{cN}\right]\right)\nonumber\\=&
	\sum_{a=1}^{N-1}\left(\delta_{cN}\sum_{b=1}^{N-1}\left(E_{ad}\delta_{bN}-\delta_{ad}E_{bN}\right)-\delta_{d<N}\left(E_{aN}\delta_{cN}-E_{cN}\delta_{aN}\right)\right)\nonumber\\=&
	\sum_{a=1}^{N-1}\left(-\delta_{cN}\delta_{ad}\sum_{b=1}^{N-1}E_{bN}-\delta_{d<N}\delta_{cN}E_{aN}\right)\nonumber\\=&
	-\left(-\delta_{cN}\delta_{d<N}-\delta_{d<N}\delta_{cN}\right)\sum_{b=1}^{N-1}E_{bN}\nonumber
	\\=&
	-2\delta_{cN}\delta_{d<N}\sum_{b=1}^{N-1}E_{bN}
\end{align}
From the third commutator on we always obtain zero. All in all we have that 
 \begin{align}\label{HT-BA}
 	\exp{\left(-\sum_{a=1}^{N-1}E_{aN}\right)}E_{cd}\exp{\left(\sum_{a=1}^{N-1}E_{aN}\right)}=E_{cd}-\left(\delta_{cN}\sum_{a=1}^{N-1}E_{ad}-\delta_{d<N}E_{cN}\right)-\delta_{cN}\delta_{d<N}\sum_{b=1}^{N-1}E_{bN}
 \end{align}
\begin{comment}
\begin{enumerate}
    \item for $A=E_{b N}$ with $b\in \{1,\ldots,N-1\}$, we obtain 
    \begin{equation}\label{HT_El1}
        \exp{(-E)}E_{b N}\exp{(E)}=E_{b N}
    \end{equation}
    because $E_{b N}$ commutes with $E$
    \item for $A=E_{bb}$ with $b \in \{1,\ldots,N-1\}$, we obtain 
    \begin{equation}\label{HT_Ell}
        \exp{(-E)}E_{b b}\exp{(E)}=E_{b b}+E_{b N}
    \end{equation}
Indeed, using \eqref{eq:comgl} we have 
    \begin{equation}
       \left[E, E_{bb}\right]=\sum_{a=1}^{N-1}\left(E_{ab}\delta_{b N}-E_{b N}\delta_{ab}\right)=-E_{bN}
    \end{equation}
    Inserting the above commutator in \eqref{HadamardFormula}, we obtain \eqref{HT_Ell}.
    \item for $A=E_{Nb}$ with $b\in \{1,\ldots,N-1\}$, we obtain 
    \begin{equation}\label{HT-E1l}
        \exp{(-E)}E_{N a}\exp{(E)}=E_{Na}+E_{NN}-\sum_{c=1}^{N-1}\left(E_{cN}+E_{ca}\right)
    \end{equation}
  %\com{Why indices have names $b, k, j, a, b$? Fix one set for the generators}

    Indeed, using \eqref{eq:comgl} we have 
    \begin{equation}
[E,E_{Na}]=\sum_{c=1}^{N-1}\left(E_{ca}\delta_{NN}-E_{NN}\delta_{ca}\right)= (\sum_{c=1}^{N-1}E_{ca} ) -E_{NN};
\end{equation}
and 
\begin{equation}
\begin{split}
\left[E,[E,E_{Na}]\right]&=\sum_{d=1}^{N-1}\sum_{c=1}^{N-1}\left[E_{dN},E_{ca}\right]-\sum_{d=1}^{N-1}\left[E_{dN},E_{NN}\right]
\\&=
\sum_{d,c=1}^{N-1}\left(E_{da}\delta_{cN}-E_{cN}\delta_{da}\right)-\sum_{d=1}^{N-1}\left(E_{dN}\delta_{NN}-E_{NN}\delta_{dN}\right)
\\=&
-2\sum_{c=1}^{N-1}E_{cN};
\end{split}
\end{equation}
therefore we have that 
\begin{equation}
	content...
\end{equation}
Inserting the above commutators in \eqref{HadamardFormula}, we obtain \eqref{HT-E1l}.
\item for $A=E_{ba}$ with $a,b \in \{1,\ldots,N-1\}$ we obtain 
\begin{equation}\label{HT-Ekl}
    \exp{(-E)}E_{a b}\exp{(E)}=E_{ba}+E_{bN}
\end{equation}
   Indeed, using \eqref{eq:comgl} we have 
\begin{equation}
[E,E_{ba}]=\sum_{c=1}^{N-1}\left(E_{ca}\delta_{b1}-E_{bN}\delta_{c a}\right)=-E_{bN};
\end{equation}
Inserting the above commutator in \eqref{HadamardFormula}, we obtain \eqref{HT-Ekl}.
\item for $A=E_{NN}$ we obtain 
\begin{equation}\label{HT-E11}
    \exp{(-E)}E_{NN}\exp{(E)}=E_{NN}-\sum_{a=1}^{N-1}E_{aN}
\end{equation}
  Indeed, using \eqref{eq:comgl} we have

@@

We multiply both sides by $R_{x}\exp{(E^{x})}$

@@

%By inserting the identity $I=\exp{(E^{x})}\exp{(-E^{x})}$ and 
By using \eqref{HT-BA} and the fact that the Casimir operator \eqref{secondCasimir} is central for the algebra we have 
then, in the right hand side

@@

% Using 
 \eqref{HT_El1}, \eqref{HT_Ell},\eqref{HT-E1l}, \eqref{HT-Ekl}, \eqref{HT-E11} we have that 
